---
title: "R TU Affordable Housing"
output: html_notebook
---


```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(httr)
library(leaflet)
library(sf)
library(lubridate)
```


Before delving into the project itself, I'll start with some exploratory data analysis.

```{r message=FALSE, warning=FALSE}
filtered_sales <- read_csv('../data/filtered_sales.csv')
lihtc <- read_csv('../data/LIHTC_updated.csv')
barnes <- read_csv('../data/barnes.csv')
property_details <- read_csv('../data/property_details.csv')

all_sale_initial_sf <- readRDS('../data/all_sale_li.rds')
all_sale_final_sf <- readRDS('../data/nearli_sales.rds')
```


Playing around with filtering and arranging (as well as referencing the website) I found out that: 

- NASHVILLE SENIOR CARE, LLC bought the most expensive property, for almost \$50 million.

-   there are 57 distinct transactions that went for the EXACT same price, \$22058426

With code below, changing the filter values: - there are 2147 sales values that appear at least 5 times

-   there are 1248 unique sales values that appear at least 10 times

-   there are 661 unique sales values that appear at least 30 times

-   there are 256 unique sales values that appear at least 100 times

    filtered_sales %>% 
      distinct(apn, .keep_all= TRUE) %>% 
      group_by(amount) %>% 
      count() %>% 
      #filter(n >= 100)
      ggplot(aes(x=amount)) +
      geom_histogram(binwidth=10000, na.rm=TRUE) +
      xlim(0,1000000)

It was concluded duplicates (same apn and date), must be errors.

First sale was made 1/1/1995, last one 10/27/2022. Most distinct sales happened in 1995, gradually going down, with an especially noticeable dip during the market crash.

    filtered_sales %>% 
      distinct(apn, .keep_all= TRUE) %>% 
      mutate(owneryear=format(ownerdate,'%Y')) %>% 
      group_by(owneryear) %>% 
      count() %>% 
      ggplot(aes(y=owneryear, weight=n)) +
      geom_bar()

    filtered_sales %>% 
      #distinct(apn, .keep_all= TRUE) %>% 
      mutate(owneryear=format(ownerdate,'%Y')) %>% 
      group_by(owneryear) %>% 
      summarize(max_price = max(amount), min_price = min(amount)) %>% 
      mutate(price_diff = max_price - min_price) %>% 
      arrange(desc(price_diff))

Biggest price difference (between min and max sale) was in 2018, no surprise there. HHowever, it's interesting to me that the fourth biggest was in 2006.

so 177 LIHTC projects, most of them placed in service in 1988 and 1989. There's 56 that were placed in service after (including) year 2000, not including the 2 placed in service unknown when and the 3 unconfirmed.

    lihtc %>% 
      filter(YR_PIS >= 2000) %>% 
      group_by(YR_PIS) %>% 
      count()

## Part 1: Statistical Analysis


1.  Using the sf library, find the closest development to each home. Hint: You can convert a tibble to an sf object using the st_as_sf function (<https://r-spatial.github.io/sf/reference/st_as_sf.html>). See, for example, this stackoverflow post: <https://gis.stackexchange.com/questions/222978/lon-lat-to-simple-features-sfg-and-sfc-in-r>. Once converted, you can use the st_nearest_feature function (<https://r-spatial.github.io/sf/reference/st_nearest_feature.html>).

> first of all, have to filter duplicate sales by doing this, removed duplicate sales removed 7507 lines, coming to 311327 observations. Next, keep only the ones that have information in property_details (it appears all of them do). Additional note: by doing a full join and filter(is.na(amount)), found out that there are 33566 values that are present in property_details but we have no info on sales.

```{r}
# note for regex in R, \\. matches for the symbol . instead of any character
# sub(".*\\s([0-9]+)\\snomination.*$", "\\1", awards)
property_sales_details <- filtered_sales %>% 
  distinct() %>% 
  inner_join(property_details, by='apn') %>% 
  mutate(
    longitude=str_extract(centroid, '-\\d*.\\d*'),
    latitude=str_remove(str_extract(centroid, ',\\d*.\\d*'),',')
  ) 

```

> Secondly, since we only want affordable housing since year 2000, filter out the dataset, going from 177 rows down to 57; but then we can add the barnes dataset to it, remove all not relevant columns for our analysis. Once that is done, there is a total of 66 rows

```{r}
barnes <- barnes %>% 
  mutate(HUD_ID=c('TNA081023A00400CO',
                  'TNA09211010400',
                  'TNA081023A90000CO',
                  'TNA08600035500',
                  'TNA10613000800',
                  'TNA05108017800',
                  'TNA09311022700',
                  'TNA06912006600',
                  'TNA10601016800'),
         .before = 'Barnes Year') %>% 
  select(HUD_ID, YR_PIS='Barnes Year', LATITUDE= lat, LONGITUDE=lng)


lihtc <- lihtc %>% 
  filter(YR_PIS >= 2000 & !YR_PIS %in% c(8888, 9999)) %>% 
  select(HUD_ID, YR_PIS, LATITUDE, LONGITUDE)

total_HUD <- bind_rows(lihtc,barnes)
```


```{r}
property_sales_details_sf <- property_sales_details %>% 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326, agr = "constant")

total_HUD_sf <- total_HUD %>% 
  st_as_sf(coords = c("LONGITUDE", "LATITUDE"), crs = 4326, agr = "constant")
```




```{r}
# start using the file the group agreed on
sales_and_HUD_sf <- property_sales_details_sf %>% 
  mutate(HUD_ID = total_HUD$HUD_ID[st_nearest_feature(property_sales_details_sf ,total_HUD_sf)]) %>% 
  left_join(total_HUD, by='HUD_ID')
```

2.  Calculate the distance from each home its closest development.

```{r}
nearest_HUD <- total_HUD_sf[st_nearest_feature(property_sales_details_sf, total_HUD_sf), ]

#note, the dist. is in m
sales_and_HUD_sf$nearest_HUD_dist <- as.numeric(st_distance(sales_and_HUD_sf, nearest_HUD, by_element = T))

#this is so we can use the file in Shiny
saveRDS(sales_and_HUD_sf, file = "../shiny_app/initial_sales_details.rds")

#and this is for the color palette in shiny
saveRDS(colorBin("Spectral",
                 domain = sales_and_HUD_sf %>% 
                  distinct(apn, .keep_all=TRUE) %>% 
                  pull(nearest_HUD_dist),
                 bins = c(0, 804.672, 1609.34, 2414.02, 3218.68, 4828.02, 6437.36, 8046.7, 9656.04, 11265.38, Inf)),
        file = "../shiny_app/pal_initial.rds"
)

```

3.  Filter the homes down to those that are within one mile of an affordable housing development.

4.  For each remaining home, calculate a new column called "group", which is defined according to the following rules. Hint: Use the case_when function to do this.

-   "pre" - for homes where the distance is less than half a mile and whose sale date was 2-5 years prior to the input year
-   "mid" - for homes where the distance is less than half a mile and whose sale date was 0-2 years prior to the input year
-   "post" - for homes where the distance is less than half a mile and whose sale date was after the input year
-   "outside" - for homes where the distance is more than half a mile and whose sale date was no more than 5 years prior to the input year
-   "other" - All other rows

> note: 1 mile = 1609.34 m - this filters it down to 112,085 values

5.  Filter out all rows whose group is "other".

> down to 72,494 rows

6.  Add an id column containing the id value for the development.

> already done that with the modifications I had to do earlier, the HUD_ID

7.  Create a column "Tpost" that, for homes in the "post" group gives the number of years that the sale occurred after the housing development was placed in service.

8.  Create a column named "age" which gives the age of the home at the time of sale.

```{r}
sales_and_HUD_sf <- sales_and_HUD_sf %>% 
  filter(as.numeric(nearest_HUD_dist) <= 1609.34) %>% 
  mutate(group=case_when(
    nearest_HUD_dist < 804.67 &
      (YR_PIS-year(ownerdate)) %in% 2:5 ~ 'pre',
    nearest_HUD_dist < 804.67 &
      (YR_PIS-year(ownerdate)) %in% 0:1 ~ 'mid',
    nearest_HUD_dist < 804.67 &
      (YR_PIS-year(ownerdate)) < 0 ~ 'post',
    nearest_HUD_dist >= 804.67 &
      (YR_PIS-year(ownerdate)) <= 5 ~ 'outside',
    TRUE ~ 'other'
  )) %>% 
  filter(group!='other') %>% 
  mutate(Tpost=if_else(group == 'post', year(ownerdate)-YR_PIS, as.numeric(NA), ),
         age=year(ownerdate)-year_built,
         land_area=as.numeric(str_remove(land_area, ' Acres'))
  ) %>% 
  filter(age >= 0 & abs(year(ownerdate)-YR_PIS) <=5)
```

9.  Filter down to only sales that took place within the five years before or after the associated development was placed in service. Then build a linear model with target variable the sales amount using the following features:

-   square_footage
-   age of home at time of sale
-   group
-   year
-   tract
-   How can you interpret the coefficients of this model?


```{r}
# the log model fits better!
#sales_and_HUD_lm <- sales_and_HUD_sf %>% 
#  filter(abs(year(ownerdate)-YR_PIS) <=5) %>% 
#  lm(amount ~ square_footage + age + factor(group) + year(ownerdate) + factor(tract), data = .)

#anova(sales_and_HUD_lm)
#summary(sales_and_HUD_lm)
```

> The intercept in this case doesn't make much sense. The square footage and year (bought) both have a positive inpact on increasing the price, as well as the group outside of the housing unit.

10. Now, try a model with target being the log of the sale price.

-   square_footage
-   age of home at time of sale
-   group
-   year
-   tract
-   How can you interpret the coefficients of this model?

```{r}
sales_and_HUD_sf %>% 
  head()
```


```{r}
sales_and_HUD_lmlog <- sales_and_HUD_sf %>% 
  lm(log(amount) ~ square_footage + age + factor(group) + year(ownerdate) + factor(tract), data = .)

sales_and_HUD_lmlogmod <- sales_and_HUD_sf %>% 
  lm(log(amount) ~ square_footage + age + land_area + factor(group) + factor(building_condition) + year(ownerdate) + factor(tract), data = .)

anova(sales_and_HUD_lmlog, sales_and_HUD_lmlogmod)
summary(sales_and_HUD_lmlogmod)
```

Bua worked on the following code:

```{r}
boarder_shapefile <- st_read("../data/Davidson County Border (GIS)/geo_export_b5e434a6-a620-409b-a8a2-610c823b7c51.shp", as_tibble = T, quiet = T) %>%
  st_transform('+proj=longlat +datum=WGS84')
```

```{r}

bins <- c(0, 0.5, 1, 1.5, 2, 3, 4, 5, 6, 7, Inf)
pal <- colorBin("Spectral", domain = all_sale_initial_sf %>% 
  distinct(apn, .keep_all=TRUE) %>% 
  pull(dist), bins = bins)

map <- leaflet(options = leafletOptions(minZoom = 10, preferCanvas = TRUE)) %>%
  addProviderTiles(provider = "CartoDB.Voyager") %>%
  setView(lng = -86.7816, lat = 36.1627, zoom = 10) %>%
  setMaxBounds(lng1 = -86.7816 + 1, 
               lat1 = 36.1627 + 1, 
               lng2 = -86.7816 - 1, 
               lat2 = 36.1627 - 1) %>%
  addCircleMarkers(data = all_sale_initial_sf %>%
                     distinct(apn, .keep_all=TRUE),
                   radius = 1.5,
                   color = ~pal(dist),
                   weight = 1,
                   fillColor = ~pal(dist),
                   fillOpacity = 0.75,
  ) %>%
  addCircleMarkers(data = total_HUD_sf,
                   radius = 2.5,
                   color = "black",
                   weight = 2,
                   fillColor = "gray",
                   fillOpacity = 0.75) %>%
  addPolylines(data = boarder_shapefile) 

map



#  addMarkers(data = sales_and_HUD_sf,
#                   clusterOptions = markerClusterOptions()
#             ) %>%

```


```{r}
# saving the final data for the app

# this was when I was saving st objects
# st_write(total_HUD_sf,'../shiny_app/initial_HUD_sf', driver='ESRI Shapefile')

# and then this to read it in shiny:
# initial_HUD_sf <- st_read(dsn='initial_HUD_sf', layer='initial_HUD_sf')

saveRDS(total_HUD_sf, file = "../shiny_app/initial_HUD_sf.rds")

saveRDS(st_drop_geometry(sales_and_HUD_sf) %>% 
  distinct(HUD_ID) %>% 
  left_join(total_HUD) %>% 
  select(HUD_ID, YR_PIS, LATITUDE, LONGITUDE) %>% 
  st_as_sf(coords = c("LONGITUDE", "LATITUDE"), crs = 4326, agr = "constant"),
  '../shiny_app/final_HUD_sf.rds')

saveRDS(sales_and_HUD_sf,'../shiny_app/final_sales_details_sf.rds')



```


```{r}
county.lines <- subset(map_data('county'),region=='tennessee' & subregion=='davidson')

gplot <- ggplot(all_sale_initial_sf)+
      geom_sf(aes(colour=dist))+
      geom_sf(data = total_HUD_sf, shape=23) +
      geom_path(data = county.lines, mapping=aes(x=long, y=lat, group=group)) +
      scale_colour_gradientn(colours=rainbow(100)[75:1])+
      labs(colour='Distance')

ggsave('initial_heatmap.png', plot=gplot)
```

```{r}
gplot2 <- ggplot(all_sale_final_sf)+
  geom_sf(aes(colour=dist))+
  geom_sf(data = total_HUD_sf, shape=23) +
  geom_path(data = county.lines, mapping=aes(x=long, y=lat, group=group)) +
  scale_colour_gradientn(colours=rainbow(100)[75:1])+
  labs(colour='Distance')

ggsave('final_heatmap.png', plot=gplot2)
```


11. Continue to explore the data to see if you can improve the models you have.
